\normalfont\spacedallcaps{Hazardous asteroids forecast via Markov random fields}

### 2021-11-30 Manuscript/manuscript.tex
PUB

\chapter{Introduction} The description and the forecast of physical \footnote{Where here physical should be intended as mesuarable} phenomena can be attacked with two different approaches: starting from a restricted set of principles/axioms, one can formulate theoretical models that provide equations that describe such phenomena. Such approach will be called here \textit{ab-initio}. On the other side, one can start from data and fit them into some general models (with fixed or not fixed number of parameters). These methods will be called here the machine learning (ML) methods. In the ab-initio case, one can fully explain the laws obtained and an overall idea of why nature works in this way. The toll that has to be paid in this case is that, since the phenomena are much more complicated than the starting principles, a consistent part of them may be excluded by the assumption made at the beginning. Furthermore, the calculation of the solution given the dynamical equation can be computationally expensive\footnote{This is the case, for instance, of the Hartree-Fock method or Density Functional Theory: such methods rewrite, with some approximations, the Schroedinger equation into a much more computationally affordable way. However, also, in this case, the solution for solids can be costly \cite{martin_2004}}. On the other side, the ML methods, since they make much more general assumptions and work directly on data, are not limited to a particular class of phenomena. Therefore they do not require the development of a specific theory as the \textit{ab-initio} one: the same model can be applied to exoplanets habitability as well to credit risk evaluation. The drawback of this approach is that the underlying mechanism by which a forecast is preferred to another one may not be interpretable. Such dichotomy between these two methods can be explained with an example taken from astronomy: at the beginning of XX century, astronomers knew that Mercury follows an elliptical orbit that, instead of being fixed, rotates around the Sun as shown in Fig. \ref{Perihelion_precession2} Such phenomenon is called \textit{Precession of the perihelion of Mercury}. The crucial point is that there was no way to explain from the Newton gravity theory such phenomenon. At this point, scholars had two paths: describe its motion with an empirical law (perhaps with an empirical modification of Newton's gravity law) or reformulate the Newton gravity from scratch. The second approach was the one followed by A.Einstein with his general relativity theory (further details are given in the A.Zee textbook \cite{zee2013einstein}). The graphical methods represent a good compromise between these two approaches: they can be used for every physical phenomenon, and, since they provide conditional dependencies/independences, they also give an elegant and intuitive representation of their mechanisms. Therefore, if the theory of the involved process is known, one can use this class of algorithms and compare its finding with the theory. In this way, one can evaluate the quality of the forecasts and the model provided by the graphical methods. The present work aims to perform such analysis for a dataset that contain the features and the hazardousness \footnote{For Earth} of asteroids. For this dataset, indeed, the relationships between the various features are known. The present work is organized as follows: first, a brief introduction about the basic theoretical concepts of the statistical methods here used will be provided, then the dataset will be described, and finally, the principal results here obtained with graphical methods will be presented as compared with other types of machine learning algorithms. Furthermore, a recap of the celestial mechanics will be provided in Appendix A. This theory explains the connections and the meaning of the different features contained in the dataset. Finally, the code used to produce the results here reported if provided in Appendix B.

@@

In this section, we are going to review the theoretical concepts
that underlie the probabilistic methods here used: these will be
exposed following the approaches of Murphy \cite{murphy2012machine} , Koller
et al. \citep{koller2009probabilistic}, Højsgaardand et al. \citep{hojsgaard2012graphical} and Russel et al. \citep{russell2010artificial}. Furthermore, a rapid overview of the main concepts of information theory will also be provided following the Cover \citep{cover2006elements}, and MacKay \citep{mackay2003information} approaches. This is because some of its concepts in the preliminary analysis of the dataset were used. Finally, we concluded this chapter with a rapid overview of algorithm interpretability, which, together with the accuracy and the
confusion matrix, is necessary for the assessment of a ML performances. On the other side, the concepts related to the celestial
mechanics here used will be described, following the Murray
approach \cite{murray1999solar} into the Appendix A.

@@

\section{Information theory}
Given an ensemble of random variables, we can quantify the amount of information that one variable contains: such quantity is called mutual information and is a crucial concept within information theory. This approach, which was implemented by Claude Shannon decades before the probabilistic modelling, represents a complementary way by which one can attack the problem of conditional dependence between random variables. Furthermore, as shown in Fig. \ref{Information_theory_connections}, this theory provides a formidable contribution to different scientific fields. Here we would provide some basic concepts of this theory, following the Cover \cite{cover2006elements} and MacKay \cite{mackay2003information} approaches, that allow defining the concept of mutual information properly. The founding concept of information theory is entropy. This quantity expresses the uncertainty of a random variable. Given a random variable $X$ with alphabet (the accessible states) $\cite{cover2006elements}$ and probability mass function $p(x)=Pr\left\lbrace X=x \rbrace\; x \in \chi$ we define the entropy of $X$ as $H(X)=-\sum_{x \in X} p(x)log p(x)$ where the logarithm has to be considered with basis 2. In analogous way the joint entropy of two random variables $(X,Y)$ with a joint distribution p(x,y) is defined as \cite{cover2006elements}:
\begin{equation}
H(X,Y)=-\sum_{x\in \chi}\sum_{y\in \mathcal{Y}}p(x,y)\log p(x,y)
\end{equation}
Furthermore, we can define also the conditional entropy as \cite{cover2006elements}:
\begin{equation}
\begin{split}
H(X|Y)=& \sum_{x\in \mathcal{X} }p(x)H(Y|X=x)\\
=& -\sum_{x\in \mathcal{X}}p(x)\sum_{y\in \mathcal{Y}}p(x,y)\log p(y|x) \\
=& -E\log p(Y|X)
\end{split}
\end{equation}
The joint entropy and the conditional entropy are related by the chain rule \cite{cover2006elements}:
\begin{equation}
H(X,Y)=H(X)+H(Y|X)
\end{equation}
Such rule can be extended to to the following from \cite{cover2006elements}:
\begin{equation}
H(X,Y|Z)=H(X|Z)+H(Y|X,Z)
\end{equation}
Given a distribution q and another distribution p, one can quantify how inefficiently the second one describes the first one using the concept of relative entropy or Kullback-Leibler distance \cite{cover2006elements,mackay2003information}:
\begin{equation}
D(p||q)=\sum p(x)\log\frac{p(x)}{q(x)}
\end{equation}
As stated by the Gibbs inequality \cite{cover2006elements,mackay2003information}:
\begin{equation}
D(p||q)\geq 0
\end{equation}
this quantity can not be negative: the entropy of a random variable associated with another cannot have a degree of uncertainty lower to the quantity it aims to describe. On these bases, we are now ready to introduce the concept of mutual information. This is defined as  \cite{cover2006elements}:
\begin{equation}
\begin{split}
I(X;Y)&=\sum\sump(x,y)\log\dfrac{p(x,y)}{p(x)p(y)}=\\
&=D(p(x,y)||p(x)p(y)) \\
&=H(X)-H(X|Y)=H(Y)-H(Y|X)
\end{split}
\end{equation}
 As for the joint distribution also in this case, we have a chain rule  \cite{cover2006elements}:
\begin{equation}
I(X_{1},X_{2},...,X_{n};Y)=\sum^{n}_{i=1}I(X_{i};Y|X_{i-1},X_{i-2},...,X_{1})
\end{equation}
The concepts here reviewed, and the relations that interconnect them can be represented as shown in Fig. \ref{Entropy_MI}. Finally, we would report the data process inequality theorem that connects the information theory with the Markov chain: if we have a Markov chain, $X\rightarrow Y \rightarrow Z$  then $I(X;Y)\geq I(X;Z) $. As for the Gibbs inequality, the underlying idea is that no clever manipulation of the data can improve the inference that can be made on them \cite{cover2006elements,mackay2003information}. Otherwise, we would have a clear violation of the second principle of thermodynamics (see for instance, the Maxwell's demon \cite{feynman2018feynman}).

@@

The asteroid dataset was retrieved from Kaggle \cite{kaggle_dataset}, which reports into a more user-friendly mode the dataset of The Center for Near-Earth Object Studies (CNEOS) \cite{cneos+nasa}, a NASA research centre. Among the 40 features that were present in the dataset,  the following cases were excluded: the redundant features (e.g. the distance quantity evaluated in miles instead of kilometres), the features connected only to the other name of the asteroid, the features connected to the name of the orbit and the one connected with the orbiting planet ( since for all it was the Earth). Thus the features obtained were therefore reduced to 22. Their description will be postponed to Appendix A since their description cannot be disjoint with celestial mechanics concepts. Their enumeration is provided in Tab. \ref{tab_features}. Since only 16 \% of the asteroids were hazardous, we considered reducing the number of the non-hazardous to improve the portion of the hazardous ones. For this purpose, we constructed a database in which the proportion hazardous/not hazardous was 1:5: thus, all the hazardous ones were included, while the not-hazardous were randomly selected \footnote{The author is aware that in principle, the proportion should be much less unbalanced, however, the tool that has to be paid for this operation is an overall reduction of the cases in the dataset that lowers the performances of the algorithms here used. The proportion here is used as a trade-off for these two contrasting facts}. Furthermore, concerning the continuous measures, the dataset was standardised and demeaned.

@@

Let us now move on to the mixed interaction model, where the hazardousness feature was also considered.  The first model was calculated with the mgm package  \cite{mgm,haslbeck2015mgm} where the nodewise regression is used \cite{meinshausen2006high}. The $k$ parameter was set equal to two, and a cross-validation (CV) with ten folds was considered. The result is reported in Fig. \ref{mgm}. From this figure, it can be seen that the features that are conditionally dependant on the Hazardous features are: the absolute magnitude, the min orbit intersection, and the eccentricity.  The theory supports all these connections. The min orbit intersection is the main parameter for the hazardous value. On the other side, the eccentricity can be tough as a parameter that describes how near the celestial body moves at the perihelion (indeed, this parameter is correctly linked with the perihelion distance). The Absolute Magnitude is also a meaningful parameter for the hazard evaluation since, if the asteroid is too small, it will be destroyed by the earth atmosphere.  Furthermore, the parameters related to the diameters are also not connected with the orbital parameters. In addition, the mean velocity is not directly related to the mean motion, but there are the orbital parameters within as expected. Thus it can be said that the model produced is almost consistent with the astronomic laws. On this basis, the performances of the model in terms of confusion matrix, ROC curve and $\phi$ parameters can be evaluated (for this purpose the Caret R package \cite{kuhn2008building,caret} was used). The confusion matrix obtained is reported in Fig. \ref{mgm_confusion}: the ROC curve is given in Fig. \ref{ROC_mgm} and it corresponds to a $\phi=0.6$. These performances will be commented on in the next section when the performances of other ML algorithms for the same dataset will be reported. Beside the mgm algorithm/package other approaches were also considered for the evaluation of the mixed interaction graphical model: the function \textit{mmod()} of gRim package \cite{hojsgaard2012graphical} and the \textit{minforest()} of gRapHD (which uses the minForest method) \cite{de2009high}. For both cases, a stepwise algorithm was used.  Their result are reported in Fig. \ref{mmod} and \ref{minforest}. It can be seen that in these models, the minimum diameter is linked with the hazardous feature, which is correct, but what is without a physical meaning is that this quantity is also linked to the features connected with the orbital parameters. A weak link with the orbital parameters is meaningful for the magnitude (as the mgm model for the mean anomaly).  The magnitude depends on the distance between the asteroids and the observer.   However, a direct link from the volume parameters (and thus from the mass) with the orbital ones is not acceptable. In principle, one can set these links as forbidden in the algorithm (blacklist), but in the author view, a model where the result is obtained without boundaries should be preferred to those obtained with a large number of boundaries. Therefore we reject the mmod and minforest ones favouring the mgm, which was obtained without boundaries.

>>>>

\chapter{Introduction} The description and the forecast of physical \footnote{Where here physical should be intended as mesuarable} phenomena can be attacked with two different approaches: starting from a restricted set of principles/axioms, one can formulate theoretical models that provide equations that describe such phenomena. Such approach will be called here \textit{ab-initio}. On the other side, one can start from data and fit them into some general models (with fixed or not fixed number of parameters). These methods will be called here the machine learning (ML) methods. In the ab-initio case, one can fully explain the laws obtained and an overall idea of why nature works in this way. The toll that has to be paid in this case is that, since the phenomena are much more complicated than the starting principles, a consistent part of them may be excluded by the assumption made at the beginning. Furthermore, the calculation of the solution given the dynamical equation can be computationally expensive\footnote{This is the case, for instance, of the Hartree-Fock method or Density Functional Theory: such methods rewrite, with some approximations, the Schroedinger equation into a much more computationally affordable way. However, also, in this case, the solution for solids can be costly \cite{martin_2004}}. On the other side, the ML methods, since they make much more general assumptions and work directly on data, are not limited to a particular class of phenomena. Therefore they do not require the development of a specific theory as the \textit{ab-initio} one: the same model can be applied to exoplanets habitability as well to credit risk evaluation. The drawback of this approach is that the underlying mechanism by which a forecast is preferred to another one may not be interpretable. Such dichotomy between these two methods can be explained with an example taken from astronomy: at the beginning of XX century, astronomers knew that Mercury follows an elliptical orbit that, instead of being fixed, rotates around the Sun as shown in Fig. \ref{Perihelion_precession2} Such phenomenon is called \textit{Precession of the perihelion of Mercury}. The crucial point is that there was no way to explain from the Newton gravity theory such phenomenon. At this point, scholars had two paths: describe its motion with an empirical law (perhaps with an empirical modification of Newton's gravity law) or reformulate the Newton gravity from scratch. The second approach was the one followed by A.Einstein with his general relativity theory (further details are given in the A.Zee textbook \cite{zee2013einstein}). The graphical methods represent a good compromise between these two approaches: they can be used for every physical phenomenon, and, since they provide conditional dependencies/independences, they also give an elegant and intuitive representation of their mechanisms. Therefore, if the theory of the involved process is known, one can use this class of algorithms and compare its finding with the theory. In this way, one can evaluate the quality of the forecasts and the model provided by the graphical methods. The present work aims to perform such analysis for a dataset that contain the features and the hazardousness \footnote{For Earth} of asteroids. For this dataset, indeed, the relationships between the various features are known. This paper is organized as follows: first, a brief introduction about the basic theoretical concepts of the statistical methods here used will be provided, then the dataset will be described, and finally, the principal results here obtained with graphical methods will be presented as compared with other types of machine learning algorithms. Furthermore, a recap of the celestial mechanics will be provided in Appendix A. This theory explains the connections and the meaning of the different features contained in the dataset. Finally, the code used to produce the results here reported if provided in Appendix B.

@@

In this section, I am going to review the theoretical concepts
that underlie the probabilistic methods here used: these will be
exposed following the approaches of Murphy \cite{murphy2012machine} , Koller
et al. \citep{koller2009probabilistic}, Højsgaardand et al. \citep{hojsgaard2012graphical} and Russel et al. \citep{russell2010artificial}. Furthermore, a rapid overview of the main concepts of information theory will also be provided following the Cover \citep{cover2006elements}, and MacKay \citep{mackay2003information} approaches. This is because some of its concepts in the preliminary analysis of the dataset were used. Finally, I concluded this chapter with a rapid overview of algorithm interpretability, which, together with the accuracy and the
confusion matrix, is necessary for the assessment of a ML performances. On the other side, the concepts related to the celestial
mechanics here used will be described, following the Murray
approach \cite{murray1999solar} into the Appendix A.

@@

\section{Information theory}
Given an ensemble of random variables, we can quantify the amount of information that one variable contains: such quantity is called mutual information and is a crucial concept within information theory. This approach, which was implemented by Claude Shannon decades before the probabilistic modelling, represents a complementary way by which one can attack the problem of conditional dependence between random variables. Furthermore, as shown in Fig. \ref{Information_theory_connections}, this theory provides a formidable contribution to different scientific fields. Here I would provide some basic concepts of this theory, following the Cover \cite{cover2006elements} and MacKay \cite{mackay2003information} approaches, that allow defining the concept of mutual information properly. The founding concept of information theory is entropy. This quantity expresses the uncertainty of a random variable. Given a random variable $X$ with alphabet (the accessible states) $\cite{cover2006elements}$ and probability mass function $p(x)=Pr\left\lbrace X=x \rbrace\; x \in \chi$ we define the entropy of $X$ as $H(X)=-\sum_{x \in X} p(x)log p(x)$ where the logarithm has to be considered with basis 2. In analogous way the joint entropy of two random variables $(X,Y)$ with a joint distribution p(x,y) is defined as \cite{cover2006elements}:
\begin{equation}
H(X,Y)=-\sum_{x\in \chi}\sum_{y\in \mathcal{Y}}p(x,y)\log p(x,y)
\end{equation}
Furthermore, we can define also the conditional entropy as \cite{cover2006elements}:
\begin{equation}
\begin{split}
H(X|Y)=& \sum_{x\in \mathcal{X} }p(x)H(Y|X=x)\\
=& -\sum_{x\in \mathcal{X}}p(x)\sum_{y\in \mathcal{Y}}p(x,y)\log p(y|x) \\
=& -E\log p(Y|X)
\end{split}
\end{equation}
The joint entropy and the conditional entropy are related by the chain rule \cite{cover2006elements}:
\begin{equation}
H(X,Y)=H(X)+H(Y|X)
\end{equation}
Such rule can be extended to to the following from \cite{cover2006elements}:
\begin{equation}
H(X,Y|Z)=H(X|Z)+H(Y|X,Z)
\end{equation}
Given a distribution q and another distribution p, one can quantify how inefficiently the second one describes the first one using the concept of relative entropy or Kullback-Leibler distance \cite{cover2006elements,mackay2003information}:
\begin{equation}
D(p||q)=\sum p(x)\log\frac{p(x)}{q(x)}
\end{equation}
As stated by the Gibbs inequality \cite{cover2006elements,mackay2003information}:
\begin{equation}
D(p||q)\geq 0
\end{equation}
this quantity can not be negative: the entropy of a random variable associated with another cannot have a degree of uncertainty lower to the quantity it aims to describe. On these bases, we are now ready to introduce the concept of mutual information. This is defined as  \cite{cover2006elements}:
\begin{equation}
\begin{split}
I(X;Y)&=\sum\sump(x,y)\log\dfrac{p(x,y)}{p(x)p(y)}=\\
&=D(p(x,y)||p(x)p(y)) \\
&=H(X)-H(X|Y)=H(Y)-H(Y|X)
\end{split}
\end{equation}
 As for the joint distribution also in this case, we have a chain rule  \cite{cover2006elements}:
\begin{equation}
I(X_{1},X_{2},...,X_{n};Y)=\sum^{n}_{i=1}I(X_{i};Y|X_{i-1},X_{i-2},...,X_{1})
\end{equation}
The concepts here reviewed, and the relations that interconnect them can be represented as shown in Fig. \ref{Entropy_MI}. Finally, we would report the data process inequality theorem that connects the information theory with the Markov chain: if we have a Markov chain, $X\rightarrow Y \rightarrow Z$  then $I(X;Y)\geq I(X;Z) $. As for the Gibbs inequality, the underlying idea is that no clever manipulation of the data can improve the inference that can be made on them \cite{cover2006elements,mackay2003information}. Otherwise, we would have a clear violation of the second principle of thermodynamics (see for instance, the Maxwell's demon \cite{feynman2018feynman}).

@@

The asteroid dataset was retrieved from Kaggle \cite{kaggle_dataset}, which reports into a more user-friendly mode the dataset of The Center for Near-Earth Object Studies (CNEOS) \cite{cneos+nasa}, a NASA research centre. Among the 40 features that were present in the dataset,  the following cases were excluded: the redundant features (e.g. the distance quantity evaluated in miles instead of kilometres), the features connected only to the other name of the asteroid, the features connected to the name of the orbit and the one connected with the orbiting planet ( since for all it was the Earth). Thus the features obtained were therefore reduced to 22. Their description will be postponed to Appendix A since their description cannot be disjoint with celestial mechanics concepts. Their enumeration is provided in Tab. \ref{tab_features}. Since only 16 \% of the asteroids were hazardous, I considered reducing the number of the non-hazardous to improve the portion of the hazardous ones. For this purpose, I constructed a database in which the proportion hazardous/not hazardous was 1:5: thus, all the hazardous ones were included, while the not-hazardous were randomly selected \footnote{The author is aware that in principle, the proportion should be much less unbalanced, however, the tool that has to be paid for this operation is an overall reduction of the cases in the dataset that lowers the performances of the algorithms here used. The proportion here is used as a trade-off for these two contrasting facts}. Furthermore, concerning the continuous measures, the dataset was standardised and demeaned.

@@

Let us now move on to the mixed interaction model, where the hazardousness feature was also considered.  The first model was calculated with the mgm package  \cite{mgm,haslbeck2015mgm} where the nodewise regression is used \cite{meinshausen2006high}. The $k$ parameter was set equal to two, and a cross-validation (CV) with ten folds was considered. The result is reported in Fig. \ref{mgm}. From this figure, it can be seen that the features that are conditionally dependant on the Hazardous features are: the absolute magnitude, the min orbit intersection, and the eccentricity.  The theory supports all these connections. The min orbit intersection is the main parameter for the hazardous value. On the other side, the eccentricity can be tough as a parameter that describes how near the celestial body moves at the perihelion (indeed, this parameter is correctly linked with the perihelion distance). The Absolute Magnitude is also a meaningful parameter for the hazard evaluation since, if the asteroid is too small, it will be destroyed by the earth atmosphere.  Furthermore, the parameters related to the diameters are also not connected with the orbital parameters. In addition, the mean velocity is not directly related to the mean motion, but there are the orbital parameters within as expected. Thus it can be said that the model produced is almost consistent with the astronomic laws. On this basis, the performances of the model in terms of confusion matrix, ROC curve and $\phi$ coefficient (also known as Matthews correlation coefficient ) can be evaluated (for this purpose the Caret R package \cite{kuhn2008building,caret} was used). The confusion matrix obtained is reported in Fig. \ref{mgm_confusion}: the ROC curve is given in Fig. \ref{ROC_mgm} and it corresponds to a $\phi=0.6$. These performances will be commented on in the next section when the performances of other ML algorithms for the same dataset will be reported. Beside the mgm algorithm/package other approaches were also considered for the evaluation of the mixed interaction graphical model: the function \textit{mmod()} of gRim package \cite{hojsgaard2012graphical} and the \textit{minforest()} of gRapHD (which uses the minForest method) \cite{de2009high}. For both cases, a stepwise algorithm was used.  Their result are reported in Fig. \ref{mmod} and \ref{minforest}. It can be seen that in these models, the minimum diameter is linked with the hazardous feature, which is correct, but what is without a physical meaning is that this quantity is also linked to the features connected with the orbital parameters. A weak link with the orbital parameters is meaningful for the magnitude (as the mgm model for the mean anomaly).  The magnitude depends on the distance between the asteroids and the observer.   However, a direct link from the volume parameters (and thus from the mass) with the orbital ones is not acceptable. In principle, one can set these links as forbidden in the algorithm (blacklist), but in the author view, a model where the result is obtained without boundaries should be preferred to those obtained with a large number of boundaries. Therefore we reject the mmod and minforest ones favouring the mgm, which was obtained without boundaries.
